---
title: "Health Data Science Report: Psychiatric Patient Hospitalisations as a Consequence of Deprivation"
author: 'Student ID: 52198593'
date: "19/11/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Activate essential packages
library(plyr) # this package needs to be loaded before dplyr to reduce potential conflicts
library(tidyverse)
library(viridis)
library(data.table)
library(readr)
library(ggplot2)
library(sf)
library(tinytex)
library(classInt)
library(sf)
library(geojsonsf)

```
## Section 1: Background: What is the aim of your visualisation? Who is your target audience, and what is the key message you are aiming to get across?

The aim of this report is to show the impact of deprivation on psychiatric patient hospitalizations across Scotland, for the most recent period for which data are available: 2017-18 to 2019-20.  
This visualization will take the form of a map with colour-coding showing the differences between the most deprived and least deprived council areas. The target audiences are members of the public who may be interested in a simple image that highlights the impact of deprivation on the need for psychiatric care in a hospital setting. Healthcare professionals and policymakers working in the domains of mental health and and indicators linked to deprivation metrics (e.g. poverty, employment, green infrastructure) may be interested in identifying areas for targeting interventions and policies for improving quality of life.

## Section 2: Data acquisition: What data do you need to produce your visualisation? Make sure you explain every variable.

To produce this map, I require data of psychiatric patient hospitalizations from each council area in Scotland, including quintile data from the Scottish Index of Multiple Deprivation. These data can be downloaded from the Scottish Public Health Observatory 2021. Also necessary is a quality shapefile of Scotland coded to include geometry linked to council area, available from https://www.spatialdata.gov.scot/geonetwork/srv/eng/catalog.search#/home.

# Psychiatric Patient Hospitalisation Data

Definition: Patients discharged from psychiatric hospitals: 3-year rolling average number and directly age-sex standardised rate per 100,000 population. All rates have been standardised against the European standard population(ESP2013) and 2011-base population estimates. Patient is selected only once per year, based on their discharge date. Values less than 10 are suppressed. No rounding. All age-groups and sexes.

Data Source: Public Health Scotland (SMR04)

Geography code: A standard code assigned to geographical areas, which in this case is council area

Numerator: Number of patients discharged from psychiatric hospitals; denominator: total population

Indicator-measure: Age-sex standardised rate per 100,000

Quintile: Based on the Scottish Index of Multiple Deprivation (SIMD), five categories of deprivation are used from 1 - most deprived (the 20% most deprived) to 5 - least deprived (the 20% least deprived). All other variables are aligned to these data, where applicable, specifically, geography code, numerator, indicator, and confidence intervals

Period: financial year from 2002/03-2019/20, 3-year aggregates

Confidence interval: Dobson's method

Last updated: April 2021

Next update: November 2021

### Read in data files and join

The first step is to acquire the data from the Scottish Public Health Observatory as .csv files (contained within the pph_csv_files folder). All time periods for all areas were initially chosen to ensure there would be sufficient data of an appropriate quality to be able to generate an interesting map.

```{r data-acquisition, echo=FALSE, include=FALSE}

# Create a collection of .csv files saved into the current directory, first by defining the working directory where the csv files are located.
mydir = "~/Documents/psych_hospital/pph_csv_files"

# Then specify the filepath and filetypes to search for, and include the full file names in the object
myfiles = list.files(path = mydir, pattern = "*.csv", full.names = TRUE)

# Inspect the object - there should be 32 files
myfiles

# Join the files using ldply from the plyr package
all_data = ldply(myfiles, read_csv)

# View the first 10 rows of data within the RMarkdown window below
all_data

# Create a tibble of the files by combining all files into one using column headers to join the data
all_data <- as_tibble(all_data)

# Save the tibble as a data file readable by R
save(all_data, file = "all_data.rda")

# Make a backup file from the amalgamated .csv files as one .csv
write_excel_csv(all_data, "all_data.csv")

```

## Section 3: Data cleaning: How do you check your data quality? What data preparation do you need to do?

Before thinking of analysis, I like to carry out various checks to make sure the data object(s) is (are) of the type I need and coded appropriately, for example, categories are factors, numeric and integer data are recognised as such.

The data in the 'period' column needs to be from the current text of "yyyy/yy to yyyy/yy financial years; 3-year aggregates" to "yyyy-yy to yyyy-yy".

```{r data-cleaning, include=FALSE}

class(all_data) # Check the type of object to make sure data are in data.frame format
names(all_data) # Look at the column names to make sure there are no errors and identify if any names need to be changed for clarity
dim(all_data) # Dimensions of the dataset - number of rows and columns
str(all_data) # Check the data structure to see if variables are categorised correctly
summary(all_data) # Generate summary statistics to check for e.g. potential anomalies, missing values, minimum and maximum values, relationship between median and mean

# Format the text in the period column to show just the year ranges without the additional text:
all_data$period <- gsub(" financial years; 3-year aggregates", "", all_data$period)

# Then replace the forward slash (/) with a hyphen (-)
all_data$period <- gsub ("/", "-", all_data$period)

# Subset data to create a smaller dataframe of all rows, and columns 2 to 8 inclusive.
all_data_subset <- all_data[, 2:8]

# Change 'period' to factor to ensure it renders correctly on graphs and charts
all_data_subset$period <- factor(all_data_subset$period)

# Change quintile to factor
all_data_subset$quintile <- factor(all_data_subset$quintile)

# Check for unique and missing values
all_data_subset$numerator %>% unique()
all_data_subset$numerator %>% table()

```

# Observations on the data
Fortunately the data were already tidy, so there was no need at this stage to carry out any restructuring or reshaping, although it will be necessary to widen the data at a later stage to calculate differences between the 1st and 5th quintiles.

1. Overall the mean is greater than the median for numerator and indicator_measure variables  
2. There are notably large differences between 3rd quartiles and max values of above, indicating presence of outlying values  

However these measures are based on the whole dataset all_data_subset, and have not been filtered by time period or geographical location. I need to dig deeper into the data to idnetify possible temporal or spatial trends that may be worth further investigation.

# Boxplots

The first set of visuals are boxplots. As I am interested in identifying variation in psychiatric patient hospitalisations attributable to deprivation, I will visualise a subset of data from quintiles 1 (most deprived) and 5 (least deprived), across all time periods.

```{r data-boxplot, echo=FALSE, fig.width=14, fig.height=12}
# Boxplot

# Call ggplot and create a subset of data from the quintiles
# Specify the levels within the quintile column with quintile %in% c("1 - most deprived", "5 - least deprived")
# Call the aesthetics to plot time period on the x axis and indicator_measure on the y axis. Boxplots are colour-coded by quintile (using the viridis palette).
# geom_boxplot(width = 0.5) alters the width of each boxplot to improve fit on the page
# Labels are added with 'labs', where title, subtitle and caption can be specified
# The viridis palette is used to fill the boxplots
# The two groups of boxplots are placed side by side using facet_wrap based on quintile order
# Additional aesthetics are managed by the theme_bw and theme() functions.
# Theme_bw creates a simple theme with limited formatting; it is important to use theme() for adjusting other elements of formatting as it has greater flexibility than theme_bw

ggplot(subset(all_data_subset, 
              quintile %in% c("1 - most deprived", "5 - least deprived")), 
       aes(x = period, y = indicator_measure, fill = quintile)) + 
  geom_boxplot(width = 0.5) +
  labs(title = "Comparison of Psychiatric Patient Hospitalisations from most deprived\nand least deprived areas of Scotland, 2002 to 2020",
       subtitle = "Age-standardised rate per 100,000",
       caption = "Source: Scottish Public Health Observatory 2021") +
  scale_fill_viridis_d(alpha = 0.5) +
  facet_wrap(quintile~.) +
  theme_bw() +
  theme(panel.background = element_blank(), # Removes heavy background formatting
        panel.border = element_blank(), # Removes the outline border around the plot
        strip.background = element_rect(fill = "white"), # Changes the background fill from the plot titles ("1 most deprived", "5 - least deprived") from gray to colour of choice
        strip.text.x = element_text(size = 12), # Changes text size of plot titles ("1 most deprived", "5 - least deprived")
        legend.position = "none", # Removes the legend
        plot.title = element_text(vjust = 1, lineheight = 1.15), # Moves the vertical alignment of the plot title and increases the spacing between the title lines
        plot.subtitle = element_text(size = 11, vjust = 1, lineheight = 1.5), # Modifies the features of the subtitle: text size, vertical adjustment and lineheight
        plot.caption = element_text(size = 10, vjust = -1, hjust = 0), # Modifies the caption attributes, including horizontal adjustment
        axis.title.x = element_blank(), # Removes the x axis title
        axis.title.y = element_blank(), # Removes the y axis title
        axis.text.x = element_text(size = 10, color = "black", angle = 90), # Modifies the x axis text
        axis.text.y = element_text(size = 10, color = "black")) # Modifies the y axis text
```
# Summary
There are higher rate of hospitalisation for people from more deprived areas than least deprived areas.  
Most deprived areas show a greater spread of datapoints than least deprived areas
There has been an overall decline in the rate of hospitalisations since the first set of aggregated data were released covering 2003-04 to 2005-06 until the final period of 2017-18 to 2019-2020.

# Histogram
Delving a little deeper into the data to identify the distribution of the hospitalisations attributable to deprivation, and identify patterns, it is required to calculate the difference between quintile 1 and quintile 5.
To do this the data need to be reshaped into wide format using the pivot_wider function, with the two levels of quintiles 1 and 5 reshaped into separate columns.

```{r data-histogram, echo=FALSE, fig.width=14, fig.height=12}

# One approach is to filter the appropriate factor levels from 'quintile'

quint_diff <- all_data_subset %>%
  filter(quintile %in% c("1 - most deprived", "5 - least deprived"))

# View the first few rows to check all is ok
head(quint_diff)

# Then reshape the data and carry out the calculation before returning the data to 'long', or tidy format:

deprivation_diff <- pivot_wider(data = quint_diff, # Create a new variable, using the filtered data from the previous step
            id_cols = c(geography_code, period), # Specify the columns to be reshaped
            names_from = quintile, # Select the names of the factor levels to be new columns
            values_from = indicator_measure) %>% # Apply the appropriate values to their quintiles
  mutate(diff = `1 - most deprived` - `5 - least deprived`) %>% # Create a new column called 'diff', subtracting the indicator_measure of quintile 5 from quintile 1
  pivot_longer(cols = c('1 - most deprived', '5 - least deprived'), # Reshape the data into tidy format by returning the quintiles to one column
               names_to = "quintile", # Restore the original name
               values_to = "indicator_measure") # Restore values to their respective levels in the quintile column

# View the first few rows to data to make sure it works
head(deprivation_diff)

# Histograms

# Call ggplot
# Use the newly calculated differences between quintiles 1 and 5 for the x axis
# geom_histogram calls the histogram function, with bins set to 20 and colour aesthetics modified to suit
# Labels are added with 'labs', where title, caption and x and y axes specified
# The groups of histograms are placed in four rows of four using facet_wrap based on time period
# Additional aesthetics are managed by the theme_bw and theme() functions as in the previous boxplots, but with some amendments based on additional text elements

ggplot(data = deprivation_diff, aes(x=diff)) +
  geom_histogram(col = "white", fill = "blue", alpha = 0.5, bins = 15) +
  labs(title = "Psychiatric Patient Hospitalisations histograms showing differences between most deprived and least deprived areas of Scotland\n2002 to 2020",
       caption = "Source: Scottish Public Health Observatory 2021",
       x = "Age-standardised rate per 100,000",
       y = "Count") +
  facet_wrap(period~.) +
  theme_bw() +
  theme(panel.background = element_blank(),
        panel.border = element_blank(),
        strip.background = element_rect(fill = "white"),
        strip.text.x = element_text(size = 12),
        strip.text.y = element_text(size = 12),
        legend.position = "none",
        plot.title = element_text(vjust = 1, lineheight = 1.15),
        plot.subtitle = element_text(size = 11, vjust = 1, lineheight = 1.5),
        plot.caption = element_text(size = 10, vjust = -1, hjust = 0),
        axis.title.x = element_text(size = 11, vjust = -0.5),
        axis.title.y = element_text(size = 11, vjust = 1),
        axis.text.x = element_text(size = 11, color = "black"),
        axis.text.y = element_text(size = 11, color = "black"))

```

## Section 4: Data Analysis: The appropriate descriptive statistics for your data.

```{r data-analysis, echo=FALSE, include=FALSE}
#summary(deprivation_diff)

#deprivation_diff %>%
#  freq_table(diff)

#descriptive <- deprivation_diff %>%
#  group_by(quintile) %>%
#  summarise(smean = mean(diff, na.rm = TRUE),
#            ssd = sd(diff, na.rm = TRUE),
#            count = n()) %>%
#  mutate(se = ssd / sqrt(count),
#         lower_ci = lower_ci(smean, se, count),
#         upper_ci = upper_ci(smean, se, count))
#
all_data_subset %>%
  group_by(period, quintile) %>%
  filter(quintile != "Total") %>%
  summarise(q1 = quantile(indicator_measure, 0.25),
            median = median(indicator_measure),
            q3 = quantile(indicator_measure, 0.75))

```

## Section 5: Communication: The data visualisation.

Mapping data can be complex due to issues of geographical projections, shapefile geometry and other embedded attributes, and correct matching of geographical coding. Fortunately, R has a package called sf - 'simple features' - that can make this task a little easier.

The sf object needs to be joined to a dataset that includes an indicator column containing geographical reference data. These two data objects are then 'merged' by specifying the appropriate columns.
Due to the way in which colours are mapped in the sf package, it can be useful to create class intervals, or breaks, of variable 'diff' to create distinct colour fills on the map instead of a colour gradient which may be harder to see if data values are close.

By using the 'diff' column calculated previously, the values will be assigned to breaks, which will then be used to colour the council areas on the map.

```{r communication-dataviz, echo=FALSE, fig.width = 9, fig.height = 9}

# Read the shapefile into R with the following code (this may not work in RMarkdown due to errors with the shapefile, but it executes from the console)

scotland_map <- read_sf(dsn = "~/Documents/Local_Authority_Boundaries_-_Scotland", "pub_las")

# Check the type of geometry - such as polygons, multipolygons
st_geometry_type(scotland_map)

# In some cases it may be necessary to alter the projects, particularly when working with many shapefiles and datasets. To do this it is essential to know the projection. Here it is OSGB 1936 / British National Grid
st_crs(scotland_map)

# Joing the shapefile to the hospitalisation data
# Create a new object and merge the data on column 'code' from the scotland_map shapefile just created and 'geography code' from the deprivation_diff dataset
map_data_depdiff <- merge(scotland_map, deprivation_diff, by.x = "code", by.y = "geography_code")

# Select the most recent set of data
map_2020 <- map_data_depdiff %>%
  filter(period == "2017-18 to 2019-20")

# Create breaks in the data using classIntervals function, and specifying the diff variable and rounding, number of breaks, and the method by which this information is to be interpreted by ggplot
# Nine breaks are to be used
breaks_qt_2020 <- classIntervals(c(min(map_2020$diff) - .0001, map_2020$diff), n = 9, style = "quantile")

# Add these breaks to the mapping object combining the shapefile and hospitalisation dataset
quantile_breaks_2020 <- mutate(map_2020, diff_cat = cut(diff, breaks_qt_2020$brks))

# View the data
quantile_breaks_2020

# Plot the data onto a map using the object just created
# Use geom_sf to plot the shapefile and use the breaks specified in previous steps to categorise the council areas and use the correct colour based on the differences between quintiles 1 and 5
# A palette of blues, useful for colour-blindness, is used to colour the areas of the map according to the breaks
# Legend labels have been changed from the breaks values to something easier to interpret
# Similar theme formatting has been applied

p1 <- ggplot(quantile_breaks_2020) +
  geom_sf(aes(fill = diff_cat)) +
  labs(title = "Comparison of Psychiatric Patient Hospitalisations in Scotland attributable to deprivation\n2017-18 to 2019-2020",
       caption = "Source: Scottish Public Health Observatory 2021",
       fill = "Deprivation categories") +
  scale_fill_brewer(palette = "Blues",
                    labels = c("1 - least deprivation", "2", "3", "4", "5", "6", "7", "8", "9 - most deprivation")) +
  theme_bw() +
  theme(panel.background = element_blank(), # Removes background colour
        panel.border = element_blank(), # Removes the borders
        panel.grid.major = element_blank(), # Removes the gridlines
        plot.title = element_text(vjust = 1, lineheight = 1.15),
        plot.subtitle = element_text(size = 11, vjust = 1, lineheight = 1.5),
        plot.caption = element_text(size = 10, vjust = -1, hjust = 0),
        axis.text.x = element_blank(), # Removes the latitude/ longitude
        axis.text.y = element_blank(), # Removes the latitude/ longitude
        axis.ticks = element_blank()) # Removes all latitude/ longitude markers

# View the map
p1
```

## Section 6: Discussion: What are the data limitations? Explain the strengths and limitations of the approach you took for your visualisation.

The main limitations of the data are that they are 1. aggregated over three-year periods and 2. standardised by age and sex.

Data aggregation can sometimes lead to interesting or important information being masked, particularly where time series are concerned. It is difficult to make reliable inferences due to the 'dilution' of effects. Due to the standardisation approach, it is impossible to determine age and sex differences in psychiatric hospitalisations. The presentation and experience of females and males at different life stages - from adolesents to senior citizens - may yield interesting differences that could be important for developing appropriate interventions and treatments.

The value of aggregation lies in its capacity to protect the identity of individuals; where sample sizes are small it may be easier to find out who a particular person is in any given year. Patient privacy is essential to maintain confidence in the healthcare system and protect individuals from further harm by information being leaked.

The strengths of a map lie in their instant visual appeal. Maps tell interesting stories about places, the people who live in those place, and can very rapidly convey a lot of information without being overwhelming. Adding further data layers and filtering options to highlight specific information or features can be highly educational and engaging, particularly for people who are comfortable engaging with digital technology.

Limitations of maps are that they are time-consuming to create, with a lot of background knowledge of projects and geographical information systems necessary to ensure they render correctly. Although there are several software packages that make map creation easier, there are many things that can go wrong which can cause delays to data projects. In this particular case, the map is 'flat' in that it only shows data from one time period. Whilst it is possible to create a map for each time period and stitch them together, the combination of colours and any changes over time would not be easy to detect and would not have the same impact.
